---
title: "DP Score plot in dppca"
description: >
  Differentially private PCA score visualization in dppca,
  including private plotting frames, bin selection, additive DP histograms, sparse DP histograms, and group-wise score histograms.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DP Score plot in dppca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```


A PCA score plot is a standard visualization tool for understanding the low-dimensional structure of high-dimensional data. 
The `dppca` package represents the private PCA score plot through a two-dimensional differentially private histogram.

## PC score plot

Let
\[
X \in \mathbb{R}^{n \times p}
\]

be the preprocessed data matrix. Typically, this means that the data have been
centered, and possibly standardized.

Let
\[
V_k = [v_1,\ldots,v_k] \in \mathbb{R}^{p \times k}
\]

be the pc direction matrix where each column $v_\ell$ is a pc direction.

The $k$-dimensional PCA score for the $i$-th observation is

\[
z_i
=
V_k^\top x_{i}
\in
\mathbb{R}^k
\quad \text{where} \quad
x_{i} \text{ denotes the } i \text{-th row of } X.
\]

In visualization, we usually select two components rather than plotting all $k$ score dimensions.

Let $(a,b)$ be the two selected component indices. A common choice is $(a,b) = (1,2)$.

For each observation, define the two-dimensional score point

\[
s_i
=
(z_{i,a}, z_{i,b})
\in
\mathbb{R}^2,
\qquad
i=1,\ldots,n.
\]

The collection $S = \{s_i\}_{i=1}^n$ is the set of two-dimensional score points.

A non-private score plot would draw all points $s_1,\ldots,s_n$ directly. 
In the private setting, we instead approximate the empirical distribution of $S$ through a private histogram.


### DP PC Score plot

In `dppca`, the private score visualization is based on the following idea.

1. Construct a private plotting frame $F \subset \mathbb{R}^2$.
2. Divide $F$ into rectangular bins.
3. Count how many score points fall into each bin.
4. Add privacy noise to the bin counts.
5. Normalize and visualize the noisy bin frequencies.


## 1. DP plotting frame

Before building a two-dimensional histogram, we need to determine the plotting region. 
This region is called the plotting frame.

If the frame is too narrow, many points may be excluded. 
If it is too wide, the histogram may become too sparse. 
Therefore, the frame should cover the relevant range of score coordinates while remaining stable under privacy constraints.

The `dppca` package constructs a square plotting frame using private quantile estimation.

### Vectorizing two-dimensional scores

Let $S \in \mathbb{R}^{n \times 2}$ be the matrix of two-dimensional scores, where the \(i\)-th row is $s_i^\top = (z_{i,a}, z_{i,b})$.

To construct one common range for both axes, we stack the two score coordinates into a single vector.

\[
s^{*}
=
(s_1^{*},\ldots,s_m^{*})^\top
\in
\mathbb{R}^m,
\qquad
m = 2n.
\]

Here, \(s^{*}\) contains all values from both selected score coordinate $\{s_{i,a}^{*}\}_{i=1}^n \cup \{s_{i,b}^{*}\}_{i=1}^n$.
This converts the problem of choosing a two-dimensional plotting range into a one-dimensional private quantile estimation problem.

### Private lower and upper quantiles

To determine the lower and upper boundaries of the plotting frame, define

\[
q_{\min} = \frac{1}{m},
\qquad
q_{\max} = \frac{m-1}{m}.
\]

These levels correspond approximately to the minimum and maximum of the stacked score coordinates.

Let $\widetilde q_{\min}$ and $\widetilde q_{\max}$ be private estimates of the $q_{\min}$ and $q_{\max}$-quantiles of the vector $s^{*}$.

In `dppca`, these private quantiles may be computed using a smooth sensitivity based DP quantile estimator in [Nissim, Raskhodnikova, and Smith (2007)](#ref-Nissim2007).

### Square plotting frame

Given the private lower and upper quantile estimates, define the center

\[
c
=
\frac{\widetilde q_{\min}+\widetilde q_{\max}}{2}
\]

and the half-length

\[
L
=
\frac{\widetilde q_{\max}-\widetilde q_{\min}}{2}.
\]

To add a visual margin and reduce boundary effects, introduce an inflate parameter $\alpha > 0$.

The inflated half-length is $L' = (1+\alpha)L$.
The final square plotting frame is

\[
F
=
[c-L',c+L']
\times
[c-L',c+L'].
\]


## 2. Choosing the number of bins

After the plotting frame $F$ has been determined, we divide it into histogram bins. 

Let $m_{\mathrm{axis}}$ be the number of bins per axis. 
Then the two-dimensional score histogram has $m = m_{\mathrm{axis}}^2$ bins in total.

The number of bins can be chosen by the user or selected using a heuristic rule.
In `dppca`, the user may specify $m_{\mathrm{axis}}$  directly.

When a default choice is needed, we use common asymptotic rules as practical guidelines. 
Following common rules of thumb, we consider

\[
m_{\mathrm{axis}} \asymp n^{1/(2+d)}
\]

as in [Wasserman and Zhou (2010)](#ref-Wasserman2010), and

\[
m_{\mathrm{axis}}
\asymp
\left(
\frac{n}{\log n}
\right)^{1/(1+d)}
\]

as in [Lei (2011)](#ref-Lei2011). Since the score histogram is two-dimensional, we take \(d=2\).


These rules are used as starting points rather than fixed prescriptions. 
In practice, the best bin choice depends on the sample size, the privacy budget, and the amount of structure in the score plot.


## 3. Make two-dimensional histogram

Let the private plotting frame be divided into bins $B_1,\ldots,B_m$.
For the score point set $S = \{s_i\}_{i=1}^n$, the non-private count in bin $B_k$ is

\[
c_k
=
\sum_{i=1}^n
\mathbf{1}\{s_i \in B_k\},
\qquad
k=1,\ldots,m.
\]

The count vector is $c = (c_1,\ldots,c_m) \in \mathbb{N}^m$.
If the histogram is normalized to a frequency vector, then

\[
q_k
=
\frac{c_k}{\sum_{j=1}^m c_j}
=
\frac{c_k}{n},
\qquad
k=1,\ldots,m
\]

The final private score visualization displays a noisy version of the frequency vector $q = (q_1,\ldots,q_m)$.



### Sensitivity of histogram counts

Under row-level adjacency, two neighboring datasets differ in one observation.
Changing one observation can move one score point from one bin to another.

Therefore, the count vector can change by at most $+1$ in one bin and $-1$ in another bin. Hence,
\[
\Delta_1(c) \leq 2 \quad \text{and} \quad \Delta_2(c) \leq \sqrt{2}
\]

These sensitivity bounds are used to privacy noise for the histogram mechanism.


### Privacy accounting

The DP score histogram procedure has two main privacy-consuming steps.

1. private quantile estimation for constructing the plotting frame,
2. private histogram release.

Let the frame construction budget be $(\epsilon_{\mathrm{frame}},\delta_{\mathrm{frame}})$
and the histogram budget be $(\epsilon_{\mathrm{hist}},\delta_{\mathrm{hist}})$.


By basic composition, the combined procedure satisfies $(\epsilon_{\mathrm{frame}}+\epsilon_{\mathrm{hist}},
\delta_{\mathrm{frame}}+\delta_{\mathrm{hist}}) \text{-DP}$.


## Method1: Additive DP histogram

A simple DP histogram can be constructed by adding independent noise to each bin count. 
The noisy counts are then post-processed to make them nonnegative and normalized. 
This additive-noise approach is commonly used in DP histogram methods [Wasserman and Zhou (2010)](#ref-Wasserman2010), 
and the procedure is summarized in [Additive DP histogram](algorithms.html#alg-add-hist). 


## Method2: Sparse DP histogram

When many bins are empty, adding noise to every bin can dominate the visualization. 
A sparse histogram aims to report only bins whose counts are large enough to be distinguishable from noise.

To address this, we use a sparse histogram procedure based on [Karwa and Vadhan (2017)](#ref-Karwa2017), 
summarized in [Sparse DP histogram](algorithms.html#alg-sparse-hist).


## Group-wise DP score histograms

When group labels are available, DP score histograms can also be constructed
separately for each group. Let
\[
\{(s_i,g_i)\}_{i=1}^n
\]
denote the score data with group labels, where \(s_i \in \mathbb{R}^2\) is the
two-dimensional PCA score and \(g_i \in \mathcal{G}\) is the group label.

For each group \(g \in \mathcal{G}\), we construct a private histogram using the
same plotting frame and the same grid. 
For each group \(g\in\mathcal{G}\), define the group-specific bin count

\[
c_k^{(g)}
=
\sum_{i=1}^n
\mathbf{1}\{s_i \in B_k,\; g_i = g\}.
\]

This gives a collection of group-wise private histograms $\{\widehat q^{(g)}\}_{g\in\mathcal{G}}$.

These histograms can be used to compare the low-dimensional PCA score structure
across groups.

In `dppca`, the group-wise version can be constructed using either the
[group-wise additive DP histogram](algorithms.html#alg-group-add-hist) or
the [group-wise sparse DP histogram](algorithms.html#alg-group-sparse-hist).


## Example usage

The exact function arguments may depend on the installed version of `dppca`.

A typical workflow is as follows.

```{r, eval = FALSE}
dp_score_plot(
  X,
  k = 2,
  eps_total = 1,
  delta_total = 1e-6,
  bin_method = "WZ"
)
```


For group-wise score histogram:

```{r, eval = FALSE}
dp_score_plot_group(
  X = gau_g,
  G = "color",
  eps_total = 5,
  delta_total = 1e-6,
  center = TRUE,
  standardize = FALSE,
  bin_method = "WZ"
)

```


## References

<span id="ref-Nissim2007"></span>
Kobbi Nissim, Sofya Raskhodnikova, and Adam Smith. (2007). "Smooth sensitivity and sampling in private data analysis". In Proceedings of the thirty-ninth annual ACM symposium on Theory of computing (STOC '07). Association for Computing Machinery, New York, NY, USA, 75–84. https://doi.org/10.1145/1250790.1250803

<span id="ref-Lei2011"></span>
Lei, Jing (2011). "Differentially private M-estimators". Advances in Neural Information Processing Systems, 24. Curran Associates, Inc.
https://proceedings.neurips.cc/paper_files/paper/2011/file/f718499c1c8cef6730f9fd03c8125cab-Paper.pdf


<span id="ref-Wasserman2010"></span>
Wasserman, L., & Zhou, S. (2010). "A Statistical Framework for Differential Privacy". Journal of the American Statistical Association, 105(489), 375–389. https://doi.org/10.1198/jasa.2009.tm08651


<span id="ref-Karwa2017"></span>
Vishesh Karwa and Salil Vadhan. (2017). "Finite sample differentially private confidence intervals". Proceedings of ITCS 2018, LIPIcs, 94, 44:1--44:9. http://drops.dagstuhl.de/opus/volltexte/2018/8344/


